---
title: "Chapter 1: Exploratory Data Analysis"
author: "Student Name"
date: today
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
    df-print: paged
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
library(eda4mlr)
```

## Learning Objectives

<!-- Review the learning objectives listed in the chapter and note them here. -->



## Reading Notes

<!-- Key concepts, questions, and observations from reading the chapter. -->



## The Big Picture

> Given a new dataset, how do you begin to understand what's in it---before fitting any models?

---

## Choosing Your Dataset

For this chapter, you'll select a dataset to explore. This mirrors real practice: EDA begins with finding data worth investigating.

**Criteria for your dataset:**

- At least 50 observations
- At least 2 variables, with at least 1 numeric variable
- Something you're genuinely interested in

**Where to look:**

- [Kaggle Datasets](https://www.kaggle.com/datasets)
- [UCI Machine Learning Repository](https://archive.ics.uci.edu/)
- [TidyTuesday](https://github.com/rfordatascience/tidytuesday)
- Government open data portals (data.gov, ONS, etc.)

**Fallback options:** If you'd prefer a ready-made dataset, `palmerpenguins::penguins` or `ggplot2::diamonds` work well.

```{r}
#| label: load-data

# Load your dataset here
# Example: my_data <- read.csv("path/to/your/data.csv")
# Example: my_data <- palmerpenguins::penguins
```

---

## Exercise 1: First Look

Before computing anything, orient yourself to the data.

### Task 1.1: Structure

Examine the dimensions and variable types. How many observations? How many variables? Which are numeric, which are categorical?

```{r}
#| label: ex1-structure

# Your code here
```

### Task 1.2: Initial Questions

Based on what you see, write down 3--5 questions you'd want to investigate. These don't need to be answerable yet---just genuine curiosities.

*Your questions:*

1.
2.
3.

### Task 1.3: Exploratory vs. Confirmatory

Which of your questions are exploratory (open-ended, pattern-seeking) and which are confirmatory (testing a specific hypothesis)? Is this distinction always clear?

*Your response:*



---

## Exercise 2: Summarizing and Visualizing

Now dig into the distributions.

### Task 2.1: Summary Statistics

Choose one or two numeric variables. For each, compute the mean, median, SD, and IQR. Based on these values, what can you infer about the shape of each distribution?

```{r}
#| label: ex2-summaries

# Your code here
```

*Your interpretation:*



### Task 2.2: Histograms

Create histograms for the same variable(s). Do the visualizations confirm your inferences from the summary statistics?

```{r}
#| label: ex2-histograms

# Your code here
```

### Task 2.3: Choosing Measures

For each variable, would you report mean/SD or median/IQR to a colleague? Why?

*Your response:*



---

## Exercise 3: Relationships and Anomalies

Look for structure across variables.

### Task 3.1: Scatter Plot

*(Skip if your dataset does not contain two numeric variables.)*

Choose two numeric variables you suspect might be related. Create a scatter plot. What pattern (if any) do you see?

```{r}
#| label: ex3-scatter

# Your code here
```

### Task 3.2: Box Plots by Group

*(Skip if not applicable, or consider creating a grouping variable by binning one of your numeric variables.)*

Choose a numeric variable and a categorical variable. Create box plots of the numeric variable, grouped by category. What do you notice?

```{r}
#| label: ex3-boxplots

# Your code here
```

### Task 3.3: Outliers and Surprises

Did you encounter any outliers or unexpected patterns? How would you investigate further?

*Your response:*



---

## Practice Problems

### Problem 1

Return to the questions you wrote in Exercise 1. Can you now answer (or partially answer) any of them? What additional data or analysis would you need?

*Your response:*



### Problem 2

Swap datasets with a classmate (or choose a second dataset yourself). Spend 10 minutes exploring it cold. What's the first thing you look at? What questions emerge?

*Your response:*



---

## Reflection

<!-- What did you learn? What remains unclear? What would you do differently next time? -->

