---
title: "Chapter 14 Workbook: Frequency Domain Methods"
subtitle: "EDA for Machine Learning"
author: "Your Name"
date: today
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
library(astsa)
library(eda4mlr)

```

## Learning Objectives

Here are the learning goals for this chapter:

1. Interpret the spectrum as the decomposition of variance across frequencies.
2. Estimate the spectral density function using the periodogram.
3. Explain why the periodogram must be smoothed to obtain a consistent estimator of the spectral density function.
4. Identify periodic components from spectral peaks and convert frequency to period.
5. Explain the bias-variance tradeoff in spectral estimation (bandwidth selection).
6. Apply coherence analysis to assess the relationship between two time series across frequencies.
7. Describe the spectral density function of an AR(1) or MA(1) model.

## The Big Picture

> Instead of asking "how does the past predict the future?", ask "what frequencies contribute to the variation?"—a complementary view of time series structure.

---

## Choosing Your Time Series

For this workbook, select a series with clear periodic structure:

**Suggested sources from `astsa` package:**

- `astsa::soi` — Southern Oscillation Index (El Niño signal)
- `astsa::rec` — recruitment (pairs well with SOI for coherence)
- `astsa::sunspotz` — sunspot numbers (strong ~11 year cycle)
- `astsa::EQ5` / `astsa::EXP6` — seismic data

```{r}
#| label: load-data

# Load your time series here
# Example: my_ts <- astsa::soi

```

---

## Exercise 1: The Periodogram

### Task 1.1: Compute the Periodogram

The periodogram decomposes the variance of a time series across frequencies. Compute it using `stats::spectrum()` or `astsa::mvspec()`.

```{r}
#| label: ex1-periodogram

# Your code here

```

*What do you see? Are there dominant frequencies?*



### Task 1.2: Identify Peaks

Find the frequency with the highest spectral power. Convert this frequency to a period (period = 1/frequency, adjusted for your time units).

```{r}
#| label: ex1-peaks

# Your code here

```

*What is the dominant period? Does this match a known cycle in your data?*



### Task 1.3: The Periodogram Is Inconsistent

The raw periodogram is a poor estimator of the true spectral density—it doesn't become more accurate as sample size increases.

Explain in your own words why smoothing is necessary.

*Your explanation:*



---

## Exercise 2: Smoothed Spectral Estimates

### Task 2.1: Apply Smoothing

Compute a smoothed spectral estimate using `astsa::mvspec()` with different spans (bandwidth parameters).

```{r}
#| label: ex2-smoothing

# Your code here

```

*How does increasing the span affect the estimate?*



### Task 2.2: Bias-Variance Tradeoff

- Wider bandwidth → smoother estimate, less variance, more bias
- Narrower bandwidth → rougher estimate, more variance, less bias

Demonstrate this tradeoff by plotting spectra with different bandwidths.

```{r}
#| label: ex2-tradeoff

# Your code here

```

*What bandwidth gives the best balance for your series?*



---

## Exercise 3: Theoretical Spectra

### Task 3.1: Spectrum of AR(1)

An AR(1) process $x_t = \phi x_{t-1} + w_t$ has spectral density:

$$f(\omega) = \frac{\sigma_w^2}{1 - 2\phi\cos(2\pi\omega) + \phi^2}$$

Simulate an AR(1) process and compare the estimated spectrum to the theoretical spectrum.

```{r}
#| label: ex3-ar1-spectrum

# Your code here

```

*How well does the estimated spectrum match the theory?*



### Task 3.2: Spectrum of MA(1)

An MA(1) process $x_t = w_t + \theta w_{t-1}$ has spectral density:

$$f(\omega) = \sigma_w^2 (1 + 2\theta\cos(2\pi\omega) + \theta^2)$$

Simulate an MA(1) process and compare the estimated and theoretical spectra.

```{r}
#| label: ex3-ma1-spectrum

# Your code here

```

*Describe the shape of the MA(1) spectrum for positive vs. negative θ.*



---

## Exercise 4: Coherence

### Task 4.1: Two Related Series

Choose two time series that might be related (e.g., `astsa::soi` and `astsa::rec`, or two economic indicators).

Plot both series and their cross-correlation function.

```{r}
#| label: ex4-two-series

# Your code here

```

*Do the series appear related? At what lag?*



### Task 4.2: Coherence Analysis

Coherence measures the correlation between two series at each frequency (like R² in regression, but frequency-specific).

Compute and plot the coherence using `astsa::mvspec()`.

```{r}
#| label: ex4-coherence

# Your code here

```

*At what frequencies are the series most strongly related?*



### Task 4.3: Interpret the Coherence

High coherence at a specific frequency means the series share a common cycle at that frequency.

*What does the coherence tell you about the relationship between your two series?*



---

## Practice Problems

### Problem 1

A colleague claims they found a 7-day cycle in their data. The spectral peak is at frequency 0.143. Verify that this corresponds to a 7-day period. What would the frequency be for a 30-day cycle?

*Your calculation:*



### Problem 2

Explain in your own words why the frequency domain and time domain perspectives are complementary. Give an example of a question better answered in each domain.

*Your response:*


